---
title: "Raster Vector Lesson"
author: "Jan Verbesselt"
date: "12 Jan 2016"
output: slidy_presentation
---

# Q&A forum

Excellent to see some action!
  
  - are there any open questions to be solved?
  - [a recent question](http://www.geo-informatie.nl/courses/grs33806/question2answer/index.php?qa=60&qa_1=cannot-bind-2-lines-together)
    
# General comments

[Ubuntu Desktop versus server](http://www.ubuntu.com/download/)

Feedback via Blackboard
  
  - Did you see the feedback on Ex 1, 2 and 3?
        
# Feedback previous lessons

Start your script, and define a e.g. in and out file name and path.
Do not change the working directory within your script or certainly not within your function

Example:

```{r, eval = FALSE}
# set input folder or file
# or ask for input folder or file
ifolder <- "/path/"
ofolder <- "/path/"
```

More info via this question: "[R user input?](http://stackoverflow.com/questions/5974967/what-is-the-correct-way-to-ask-for-user-input-in-an-r-program)"

> [R examples](http://www.rexamples.com/4/Reading%20user%20input)

# Solution of Wednesday's shell script

```{r, engine='bash', results='hide', eval =FALSE}
#!/bin/bash
cd ../data
fnin="ndvi.tif"
fnout="ndvi_latlong_60.tif"
fntemp="temp.tif"
fnpng="ndvilatlong.png"
echo "resample the inputfile: $fnin"
gdalwarp -tr 60 60 -r cubic $fnin $fntemp
echo "reproject to lat/long: $fnout"
gdalwarp -t_srs EPSG:4326 $fntemp $fnout
rm $fntemp
gdal_translate -of png $fnout $fnpng
echo "The output file: $fnout"
```

> What is the benefit of the Shell script? Why would you use it?

# Tutorial of today

Questions: 

Would you rather reproject a raster or a vector layer?
    
Why can't we simply use the LineLength() function from the sp package?
    
> Hint, look at the CRS in which we are working, and at the help page of the distHaversine() function. Also the see distVincentyEllipsoid() function. Does this one provide a more accurate distance?

# Todays lesson

[Vector & Raster](http://geoscripting-wur.github.io/VectorRaster/)




